---
title: "Exploratory Analysis of Eukaryotic Genome Reference Catalog"
author: "Mikkel Winther Pedersen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8, fig.height = 5)
library(readr)
library(ggplot2)
library(gghighlight)
library(scales)
library(tidyverse)
library(dplyr)
library(inflection)
library(scales)


```

```{r load-data, echo=TRUE}
# Load the metadata file containing eukaryotic genome assemblies
wgs_euks <- read_delim("wgs_eukaryota.metadata_filter.short.tsv",
                       delim = "\t", escape_double = FALSE, trim_ws = TRUE)

iridian_genomes <- read_csv("iridiantaxid.txt", col_names = FALSE)
colnames(iridian_genomes) <- c("tax_id")


colnames(wgs_euks)

# Assigning appropriate column names
colnames(wgs_euks) <- c(
  "assembly_accession", "bioproject_accession", "biosample_description",
  "tax_id", "source_db", "biosample_updated", "biosample_owner",
  "biosample_location", "biosample_isolation", "assembly_level", "contig_l50",
  "contig_n50", "gaps_between_scaffolds", "gc_count", "gc_percent",
  "genome_coverage", "component_sequences", "contigs", "organelles",
  "scaffolds", "scaffold_l50", "scaffold_n50", "chromosomes",
  "sequence_length", "ungapped_length"
)

wgs_euks$is_iridian <- wgs_euks$tax_id %in% iridian_genomes$tax_id
# check how many are iridian genomes
table(wgs_euks$is_iridian)
```



```{r preview-db, echo=TRUE}
# Preview the unique databases included, and rename the source_db's 
unique(wgs_euks$source_db)
wgs_euks$source_db <- gsub("SOURCE_DATABASE_", "", wgs_euks$source_db)
wgs_euks$genome_coverage <- as.numeric(gsub("x", "", wgs_euks$genome_coverage))

```

### In the terminal/on server

``` bash
conda install -c bioconda taxonkit
wget -c https://ftp.ncbi.nih.gov/pub/taxonomy/taxdump.tar.gz 
tar -zxvf taxdump.tar.gz

mkdir -p $HOME/.taxonkit
cp names.dmp nodes.dmp delnodes.dmp merged.dmp $HOME/.taxonkit

cut -f4 wgs_eukaryota.metadata_filter.short.tsv | tail -n+2 | taxonkit reformat2 - > taxon_paths.tsv
```

```{r join tables, echo=TRUE}

taxon_paths <- read_delim("taxon_paths.tsv", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)
colnames(taxon_paths) <- c("tax_id", "taxonomic_path")
 
wgs_euks2 <- wgs_euks |> inner_join(taxon_paths)
  
### Lets plot the reference metadata out first

# Bar chart of genome counts by source database
ggplot(wgs_euks2, aes(x = source_db)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3, size = 4) +
  labs(title = "", x = "", y = "Number of Genomes") +
  theme_minimal() 
```

```{r barplot-stacked, echo=TRUE}
# Stacked bar chart of assembly levels within each source database
ggplot(wgs_euks2, aes(x = source_db, fill = assembly_level)) +
  geom_bar(position = "stack") +
  labs(title = "", x = "Source Database", y = "Number of Genomes", fill = "Assembly Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r barplot-proportional, echo=TRUE}
# Proportional stacked bar chart of assembly levels per source
ggplot(wgs_euks2, aes(x = source_db, fill = assembly_level)) +
  geom_bar(position = "fill") +
  labs(title = "", x = "Source Database", y = "Proportion of Genomes", fill = "Assembly Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r length_ratio, echo=TRUE}
# Step 1: Calculate length_ratio 
wgs_euks <- wgs_euks %>%
  mutate(len_ratio = ifelse(
    is.na(ungapped_length) | is.na(sequence_length) | sequence_length == 0,
    NA_real_,
    ungapped_length / sequence_length
  ))

# Step 2: Density plot colored by assembly_level and faceted by source_db
ggplot(wgs_euks, aes(x = assembly_level, y = len_ratio, fill = assembly_level)) +
    geom_violin(alpha = 0.5, na.rm = TRUE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
    facet_wrap(~ source_db) +
    scale_y_continuous() +
    labs(
        title = "Length Ratio Distribution by Assembly Level",
        x = "Assembly Level",
        y = "Length Ratio"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```


```{r scatter-n50-l50, echo=TRUE}
# Scatter plot: Scaffold N50 vs Scaffold L50, faceted by source
ggplot(wgs_euks2, aes(x = scaffold_n50, y = scaffold_l50)) +
  geom_point(aes(size = contigs), alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(title = "", x = "Scaffold N50", y = "Scaffold L50") +
  theme_minimal()
```

```{r scatter-n50-contigs-gc, echo=TRUE}
# Scatter plot: Scaffold N50 vs Contigs, filled by GC content
ggplot(wgs_euks2, aes(x = scaffold_n50, y = contigs)) +
  geom_point(aes(size = contigs, fill = gc_percent), shape = 21, color = "black", alpha = 0.8) +
  facet_wrap(~ source_db) +
  labs(title = "", x = "Scaffold N50", y = "Number of contigs", fill = "GC Percent") +
  theme_minimal()
```
## 
```{r genomes_lost_contigs, echo=TRUE}

# Define thresholds (descending from 500,000 in steps of 10,000)
thresholds <- seq(500000, 0, by = -100)

# For each source_db, count genomes <= threshold
threshold_by_source <- expand.grid(
  threshold = thresholds,
  source_db = unique(wgs_euks2$source_db)
) %>%
  rowwise() %>%
  mutate(n_remaining = sum(
    wgs_euks2$source_db == source_db &
      wgs_euks2$contigs <= threshold,
    na.rm = TRUE
  )) %>%
  ungroup()

# Make a plot by split databses
ggplot(threshold_by_source, aes(x = threshold, y = n_remaining, color = source_db)) +
  geom_point(size = 1) +
  scale_x_reverse(breaks = seq(0, 500000, 50000), labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Genome Retention by Contig Threshold, Split by Source DB",
    x = "Maximum Number of Contigs (Threshold)",
    y = "Number of Genomes Retained",
    color = "Source DB"
  ) +
  theme_minimal()


# Lets then merge the two sources and for each threshold, count genomes with contigs <= threshold
threshold_counts <- data.frame(threshold = thresholds) %>%
  rowwise() %>%
  mutate(n_remaining = sum(wgs_euks2$contigs <= threshold, na.rm = TRUE)) %>%
  ungroup()

# Reverse x and y for proper elbow detection
x <- rev(threshold_counts$threshold)
y <- rev(threshold_counts$n_remaining)

# Detect elbow (tipping point)
elbow_index <- inflection::findiplist(x, y, index = 1)
tipping_threshold <- x[elbow_index]
n_at_tipping <- y[elbow_index]

# Plot with tipping point highlighted
ggplot(threshold_counts, aes(x = threshold, y = n_remaining)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line(color = "steelblue") +
  geom_vline(xintercept = tipping_threshold, linetype = "dashed", color = "red") +
  geom_text(
    data = data.frame(
      threshold = tipping_threshold,
      n_remaining = n_at_tipping,
      label = paste0("Tipping Point: ", tipping_threshold, " contigs")
    ),
    aes(x = threshold, y = n_remaining, label = label),
    inherit.aes = FALSE,
    vjust = -1, hjust = 1.1, color = "red"
  ) +
  scale_x_reverse(breaks = seq(0, 500000, 50000), labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Tipping Point in Genome Retention Curve",
    x = "Contig Threshold",
    y = "Number of Genomes Retained"
  ) +
  theme_minimal()

elbows <- inflection::findiplist(x, y, index = 1)
tipping_thresholds <- x[elbows]

con_thres <- max(tipping_thresholds, na.rm = TRUE)


```

```{r density-gc, echo=TRUE}
# Density plot of GC percent by assembly level
ggplot(wgs_euks2, aes(x = as.numeric(gc_percent), fill = assembly_level)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ source_db) +
  labs(title = "", x = "GC Percent", y = "Density", fill = "Assembly Level") +
  theme_minimal()
```

```{r ungapped-vs-total, echo=TRUE}
# Scatter plot comparing ungapped and total sequence lengths
ggplot(wgs_euks2, aes(x = ungapped_length, y = sequence_length)) +
  geom_point(alpha = 0.8, aes(color = assembly_level)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
  facet_wrap(~ source_db) +
  labs(title = "", x = "Ungapped Sequence Length", y = "Total Sequence Length") +
  theme_minimal()

```


```{r ungapped-vs-total-ratio, echo=TRUE}
# how many sequence are 10% longer than their ungapped sequence length?

wgs_euks2 %>% filter(source_db == "GENBANK") |> 
  filter(sequence_length / ungapped_length  > 1.1) %>% 
  nrow()

# Let's add a length_ratio to the dataframe
wgs_euks2 <- wgs_euks2 %>%
  mutate(len_ratio = (sequence_length / ungapped_length))

summary(wgs_euks2$len_ratio)

len_ratio <- 1.1 
```

### plots the points filtered by the len_ratio threshold

```{r plot-len_ratio-boxplot, echo=TRUE}
# Step 2: Density plot colored by assembly_level and faceted by source_db
ggplot(wgs_euks2, aes(x = assembly_level, y = len_ratio, fill = assembly_level)) +
    geom_violin(alpha = 0.5, na.rm = TRUE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
    geom_jitter() +
    facet_wrap(~ source_db) +
    scale_y_continuous() +
    labs(
        title = "Length Ratio Distribution by Assembly Level",
        x = "Assembly Level",
        y = "How much longer is the total sequence length vs the ungapped length"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r plot-len_ratio-filter, echo=TRUE}
ggplot(wgs_euks2, aes(x = ungapped_length, y = sequence_length)) +
  geom_point(alpha = 0.8, aes(color = assembly_level)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
  facet_wrap(~ source_db, drop = FALSE) +  # This keeps all source_db panels
  gghighlight(len_ratio > 1.1, use_direct_label = FALSE) +
  labs(
    title = "Ungapped vs Total Sequence Length (Highlighting len_ratio > 1.1)",
    x = "Ungapped Sequence Length",
    y = "Total Sequence Length"
  ) +
  theme_minimal()

```

### plots the points filtered by the len_ratio threshold and that are from the iridian company
```{r plot-len_ratio-filter-iridian, echo=TRUE}
ggplot(wgs_euks2, aes(x = ungapped_length, y = sequence_length)) +
  geom_point(alpha = 0.8, aes(color = assembly_level)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
  facet_wrap(~ source_db, drop = FALSE) +
  gghighlight(is_iridian == TRUE, use_direct_label = FALSE) +
  labs(
    title = "Ungapped vs Total Sequence Length (Highlighting Iridian Genomes)",
    x = "Ungapped Sequence Length",
    y = "Total Sequence Length"
  ) +
  theme_minimal()

```


### We found quite a bit of NAs in the scaffold_n50, lets see which assembly level they are

#write.csv(wgs_euks2, file ="wgs_euks2.csv")

```{r counting-nas, echo=TRUE}
# Count NA values per column and source_db
na_counts <- wgs_euks2 %>%
  select(source_db, scaffold_n50, contigs) %>%
  pivot_longer(cols = c(scaffold_n50, contigs), names_to = "variable", values_to = "value") %>%
  mutate(is_na = is.na(value)) %>%
  group_by(source_db, variable) %>%
  summarise(n = sum(is_na), .groups = "drop") %>%
  complete(source_db, variable, fill = list(n = 0))

# Plot grouped by source_db, faceted by column
ggplot(na_counts, aes(x = source_db, y = n, fill = source_db)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Number of NAs by Column and Source DB",
       x = "Source DB",
       y = "Number of NAs") +
  theme_minimal()


# Count NAs in scaffold_n50 and show their assembly_levels
na_pie_data <- wgs_euks2   %>%
  filter(is.na(scaffold_n50)) %>%
  count(assembly_level) %>%
  mutate(percentage = n / sum(n),
         label = paste0(assembly_level, " (", round(percentage * 100, 1), "%)"))

ggplot(na_pie_data, aes(x = "", y = n, fill = assembly_level)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Distribution of NAs in scaffold_n50 by Assembly Level",
       x = NULL, y = NULL, fill = "Assembly Level") +
  theme_void()
```

Doing a manual check on these the genomes looks ok, one option could be to filter these based on the coverage they have. We observe a suspisious decline in the lower end of the coverage. This high value is caused by the extreme high values and hence skewness in the data. We therefore selected a hard thresholds of 10. For setting the contigs we use the .01 / .99 percentile of RefSeq stats to filter the Genbank genomes. as well as the length_ratio


One idea is to use the RefSeq as a gold standard, and use the scaffold_n50 and contigs statistics to filter the genbank.

```{r scaffols_n50_NA_contig_genome_coverage, echo=TRUE}

# Setting thresholds
scaf_thres <- quantile(wgs_euks$scaffold_n50[wgs_euks$source_db == "REFSEQ"], 0.01, na.rm = TRUE)
#con_thres <- quantile(wgs_euks$contigs[wgs_euks$source_db == "REFSEQ"], 0.99, na.rm = TRUE) # Is now set as the tipping point from the above elbow plot
gen_cov_thres <- 10
len_thres <- 1.1

# Count how many will be filtered by this 
highlighted_counts <- wgs_euks2 %>%
  filter(
    !(
      (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
      (is.na(scaffold_n50)    | scaffold_n50 > scaf_thres) &
      (is.na(contigs)         | contigs < con_thres) &
      (is.na(len_ratio)         | len_ratio < len_thres)
    )
  ) %>%
  group_by(source_db) %>%
  summarise(count = n())


ggplot(wgs_euks2, aes(x = genome_coverage, y = contigs)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  scale_x_log10() +
  labs(
    title = "",
    x = "genome_coverage",
    y = "contigs"
  ) +
  gghighlight(scaffold_n50 > scaf_thres, contigs < con_thres, genome_coverage >= gen_cov_thres, use_direct_label = FALSE) +
  geom_text(
    data = highlighted_counts,
    aes(x = Inf, y = Inf, label = paste("To be filtered:", count)),
    hjust = 1.1, vjust = 1.5,
    inherit.aes = FALSE,
    size = 4,
    fontface = "bold",
    color = "darkgrey"
  ) +
  theme_minimal()
```

```{r scatter-contigs-n50, echo=TRUE}
# Contigs vs Scaffold N50
# Filter input for plotting (remove NA and zero values for valid plotting)
plot_data <- wgs_euks2 %>%
  filter(!is.na(scaffold_n50), !is.na(contigs), scaffold_n50 > 0, contigs > 0)

# Step 1: Filter and count per source_db (MATCH gghighlight thresholds)
filtered_counts <- wgs_euks2 %>% filter(
    (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
    (is.na(scaffold_n50) | scaffold_n50 > scaf_thres) &
    (is.na(contigs) | contigs < con_thres) &
      (is.na(len_ratio)         | len_ratio < len_thres)
  ) %>%
  count(source_db, name = "filtered_n")

# Step 2: Compute label positions (adjust x/y to avoid overlap)
label_positions <- plot_data %>%
  group_by(source_db) %>%
  summarize(
    x = quantile(contigs, 0.05, na.rm = TRUE),
    y = max(scaffold_n50, na.rm = TRUE) * 0.95,
    .groups = "drop"
  )

# Step 3: Join label data
label_data <- left_join(filtered_counts, label_positions, by = "source_db")

# Step 4: Create the plot
ggplot(plot_data, aes(x = contigs, y = scaffold_n50)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(
    title = "Filtered Genome Count per Source DB",
    x = "Number of Contigs",
    y = "Scaffold N50"
  ) +
  theme_minimal() +
  gghighlight(scaffold_n50 > scaf_thres, contigs < con_thres,genome_coverage >= gen_cov_thres, use_direct_label = FALSE) +
  scale_x_continuous(labels = scales::label_comma()) +
  geom_text(
    data = label_data,
    aes(x = x, y = y, label = paste("Kept genomes:", scales::comma(filtered_n))),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0,
    fontface = "bold"
  )
```

```{r}

target_groups <- c(
  "Metazoa", "Fungi", "Choanoflagellates", "Nucleariids",
  "Tubulinea", "Discosea", "Mycetozoa",
  "Archaeplastida", "Chloroplastida", "Rhodophyta", "Glaucophyta",
  "Diatoms", "Brown algae", "Oomycetes",
  "Ciliates", "Dinoflagellates", "Apicomplexans",
  "Foraminifera", "Radiolarians", "Cercozoans",
  "Euglenozoa", "Fornicata", "Parabasalia", "Heterolobosea",
  "Collodictyonids", "Rigifilida", "Mantamonas",
  "Haptophytes", "Cryptophytes", "Hemimastigophora", "Ancyromonads"
)

filtered_data <- wgs_euks2 %>%
  filter(
    (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
    (is.na(scaffold_n50) | scaffold_n50 > scaf_thres) &
    (is.na(contigs) | contigs < con_thres) &
      (is.na(len_ratio)         | len_ratio < len_thres)
  )

library(stringr)
filtered_data <- filtered_data %>%
  mutate(
    matched_group = str_extract(taxonomic_path, str_c(target_groups, collapse = "|"))
  ) %>%
  filter(!is.na(matched_group))


library(dplyr)
library(ggplot2)

# Step 1: Add a filtered-in flag according to the filtering criteria
"Keep genomes that have good enough coverage (≥10), are not too fragmented, and don’t have an absurd number of contigs — but allow missing values."

wgs_euks2_flagged <- wgs_euks2 %>%
  mutate(
    keep = (
      (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
      (is.na(scaffold_n50) | scaffold_n50 > scaf_thres) &
      (is.na(contigs) | contigs < con_thres) &
      (is.na(len_ratio) | len_ratio < len_thres)
    )
  )


library(purrr)

# Step 1: Prepare pie data
pie_data <- wgs_euks2_flagged %>%
  mutate(filter_status = ifelse(keep, "Filtered In", "Filtered Out")) %>%
  count(source_db, filter_status) %>%
  group_by(source_db) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label = paste0(filter_status, "\n", percent, "%")
  ) %>%
  ungroup()

# Split data by source_db
split_pie_data <- split(pie_data, pie_data$source_db)

# Create a named list of ggplot objects
pie_plots <- map(split_pie_data, function(df) {
  ggplot(df, aes(x = "", y = n, fill = filter_status)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
    scale_fill_manual(values = c("Filtered In" = "steelblue", "Filtered Out" = "grey70")) +
    labs(title = paste("Filter Status for", unique(df$source_db))) +
    theme_void()
})

library(patchwork)
wrap_plots(pie_plots)  # Combine them all in a grid


library(forcats)

# Step 1: Split taxonomic path and trim
wgs_euks2_flagged <- wgs_euks2_flagged %>%
  separate(
    taxonomic_path,
    into = paste0("rank", 1:7),
    sep = ";",
    fill = "right"
  ) %>%
  mutate(across(starts_with("rank"), ~trimws(.)))

# Step 2: Summarize data including assembly_level
remove_data <- wgs_euks2_flagged %>%
  filter(!keep) %>%
  count(rank2, source_db, assembly_level, name = "n") %>%
  mutate(rank2 = fct_reorder(rank2, n))

# Step 3: Plot with stacked bars by assembly_level
ggplot(remove_data, aes(x = rank2, y = n, fill = assembly_level)) +
  geom_col() +
  facet_wrap(~ source_db, scales = "free_y") +
  labs(
    title = "Filtered Out Eukaryotic Groups by Source DB and Assembly Level",
    x = "Eukaryotic Group",
    y = "Number of Genomes",
    fill = "Assembly Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# lets have a more in depth look at who is removed of genbank genomes
genbank_remove <- wgs_euks2_flagged %>% filter(source_db == "GENBANK" | source_db == "REFSEQ", keep == "FALSE")
genbank_keep <- wgs_euks2_flagged %>% filter(source_db == "GENBANK", keep == "TRUE")

# lets have a more in depth look at who is removed of genbank and refseq genomes
all_remove <- wgs_euks2_flagged %>% filter(source_db == "GENBANK" | source_db == "REFSEQ", keep == "FALSE")
all_keep <- wgs_euks2_flagged %>% filter(source_db == "GENBANK" | source_db == "REFSEQ", keep == "TRUE")

write.csv(all_remove, file = "removed_genomes.csv")
write.csv(all_keep, file = "kept_genomes.csv")

# how many are iridian genomes
table(genbank_keep$is_iridian)


```

```{r scatter-contigs-n50-adding-iridians, echo=TRUE}
# Contigs vs Scaffold N50 with iridians

ggplot(genbank_keep, aes(x = contigs, y = scaffold_n50, shape = is_iridian)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(
    title = "Highlighting Iridian Genomes",
    x = "Number of Contigs",
    y = "Scaffold N50",
    shape = "Iridian Genome"
  ) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
  scale_x_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  gghighlight(is_iridian == TRUE, use_direct_label = FALSE)
```
```{r}
unique(genbank_keep$rank2)
```
```{r scatter-contigs-n50-animals, echo=TRUE}
# Contigs vs Scaffold N50 with iridians

genbank_keep |> filter(rank2 == "Chordata") |>
ggplot(aes(x = contigs, y = scaffold_n50, shape = is_iridian)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(
    title = "Plotting Chordata (animal) Genomes",
    x = "Number of Contigs",
    y = "Scaffold N50",
    shape = "Iridian Genome"
  ) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
  scale_x_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  gghighlight(is_iridian == TRUE, use_direct_label = FALSE)
```
```{r scatter-contigs-n50-plants, echo=TRUE}
# Contigs vs Scaffold N50 with iridians

genbank_keep |> filter(rank2 == "Streptophyta") |>
ggplot(aes(x = contigs, y = scaffold_n50, shape = is_iridian)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(
    title = "Plotting Streptophyta (plant) Genomes",
    x = "Number of Contigs",
    y = "Scaffold N50",
    shape = "Iridian Genome"
  ) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
  scale_x_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  gghighlight(is_iridian == TRUE, use_direct_label = FALSE)
```
  
  


```{r}

```

