---
title: "Exploratory Analysis of Eukaryotic Genome Reference Catalog"
author: "Mikkel Winther Pedersen and added to by E. Elisa Davis"
date: "`r Sys.Date()`"
output: html_document
---
#### what is Needed before this .RMD?
#1. Download Genbank + RefSeq metadata (script in Rosalind RefSeq2026) --> leading to eukaryote_master_database.tsv, this is a large metadata file
#2. Have completed your local area "whitelist" --> list of taxa found in region with associated linages + Genbank Tax_ids
#3. with taxonKit, created a taxon_path.tsv to attach lineages to eukaryote_master_database --> make taxonomic visualisation easier.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8, fig.height = 5)
library(readr)
library(ggplot2)
library(gghighlight)
library(scales)
library(tidyverse)
library(dplyr)
library(inflection)
library(scales)
library(stringr)


```

```{r load-data, echo=TRUE}
# Load the metadata file containing eukaryotic genome assemblies
wgs_euks <- read_delim("/Users/davisee/OneDrive - University of Tasmania/Documents/chapter2/raw data/eukaryote_master_database.tsv",
                       delim = "\t", escape_double = FALSE, trim_ws = TRUE)

so_RAMS_list <- read_delim("C:/Users/davisee/OneDrive - University of Tasmania/Documents/chapter2/raw data/RAMS_alleukaryote_taxlist_20260224/WoRMS_RAMS_2026-02-01/RAMS_SO_whitelist_complete.tsv", 
                           delim = "\t", escape_double = FALSE, trim_ws = TRUE)
clean_SO_RAMS_list <- so_RAMS_list %>%
  filter(!is.na(tax_id))
so_genomes <- clean_SO_RAMS_list[,2]
colnames(so_genomes) <- c("tax_id")


colnames(wgs_euks)

# Assigning appropriate column names
colnames(wgs_euks) <- c(
  "assembly_accession", "bioproject_accession","biosample_common_name", "biosample_description",
  "tax_id", "organism_name", "source_db", "biosample_owner",
  "biosample_host","biosample_location", "biosample_isolation", "biosample_ecotype",
  "assembly_level", "contig_l50", "contig_n50", "gaps_between_scaffolds",
  "atgc_count", "gc_percent", "genome_coverage", "component_sequences", "contigs", "organelles",
  "scaffolds", "scaffold_l50", "scaffold_n50", "chromosomes",
  "sequence_length", "ungapped_length", "organelle_description", "organelle_sequence_length"
)

wgs_euks$is_SO <- wgs_euks$tax_id %in% so_genomes$tax_id
# check how many are iridian genomes
table(wgs_euks$is_SO)
so_taxa <- wgs_euks %>% filter(is_SO == TRUE) %>% pull(tax_id) %>% n_distinct()
print(so_taxa)
noso_taxa <- wgs_euks %>% filter(is_SO == FALSE) %>% pull(tax_id) %>% n_distinct()
print(noso_taxa)
so_taxa / (so_taxa + noso_taxa)

```



```{r preview-db, echo=TRUE}
# Preview the unique databases included, and rename the source_db's 
unique(wgs_euks$source_db)
wgs_euks$source_db <- gsub("SOURCE_DATABASE_", "", wgs_euks$source_db)
wgs_euks_review <- wgs_euks[wgs_euks$source_db == "Source Database", ]
#remove extraneous rows that have no information
wgs_euks <- wgs_euks[wgs_euks$source_db != "Source Database", ]
unique(wgs_euks$source_db)
#finish clean up
wgs_euks$genome_coverage <- as.numeric(gsub("x", "", wgs_euks$genome_coverage))
wgs_euks$tax_id <- as.character(wgs_euks$tax_id)

```
```{r rams-taxonomy-exploration, echo=TRUE}
# Exploratory summary of phylum/class diversity in RAMS SO whitelist
# Run this first to inform target group binning decisions

so_RAMS_list %>%
  distinct(tax_id, phylum, class) %>%
  count(phylum, class) %>%
  arrange(desc(n)) %>%
  print(n = Inf)
```

```{r rams-target-groups, echo=TRUE}
# Assign target groups based on phylum and class
# Ecologically important SO groups kept separate, remainder collapsed

so_RAMS_list <- so_RAMS_list %>%
  mutate(
    target_group = case_when(
      # Vertebrates
      phylum == "Chordata" & class %in% c("Actinopteri",
                                           "Chondrichthyes",
                                           "Hyperoartia",
                                           "Myxini")            ~ "Fish",
      phylum == "Chordata" & class == "Mammalia"                 ~ "Mammalia",
      phylum == "Chordata" & class == "Aves"                     ~ "Birds",
      phylum == "Chordata" & class %in% c("Ascidiacea",
                                           "Thaliacea",
                                           "Appendicularia")     ~ "Tunicata",
      phylum == "Chordata"                                        ~ "Chordata_Other",

      # Major SO invertebrate phyla
      phylum == "Arthropoda" & class %in% c("Malacostraca",
                                           	"Ostracoda", 
                                           "Hexactinellida") ~ "Crustaceans_Artropoda",
      phylum == "Arthropoda" & class =="Pycnogonida"  ~ "SeaSpiders_Arthropoda",
      phylum == "Mollusca"                                        ~ "Mollusca",
      phylum == "Echinodermata"                                   ~ "Echinodermata",
      phylum == "Annelida"                                        ~ "Annelida",
      phylum == "Porifera"                                        ~ "Porifera",
      phylum == "Cnidaria"                                        ~ "Cnidaria",
      phylum == "Bryozoa"                                         ~ "Bryozoa",
      phylum == "Nematoda"                                        ~ "Nematoda",
      phylum == "Platyhelminthes"                                 ~ "Platyhelminthes",

      # Important SO protist groups
      phylum == "Foraminifera"                                    ~ "Foraminifera",
      phylum == "Bacillariophyta"                                 ~ "Diatoms",
      phylum == "Ciliophora"                                      ~ "Ciliophora",

      # Smaller but real SO groups collapsed
      phylum %in% c("Chaetognatha", "Brachiopoda", "Nemertea",
                    "Kinorhyncha", "Acanthocephala",
                    "Ctenophora", "Hemichordata")                 ~ "Minor_Invertebrates",

      # Fungi and plants
      phylum %in% c("Ascomycota", "Basidiomycota")               ~ "Fungi",
      phylum == "Streptophyta"  & class !%in% ("Magno")                                  ~ "Plants",

      TRUE                                                        ~ "Other"
    )
  )

cat("RAMS target group breakdown:\n")
print(table(so_RAMS_list$target_group, useNA = "always"))
```

```{r rams-target-groups-barplot, echo=TRUE}
# Bar chart — unique taxa per target group

so_RAMS_list %>%
  distinct(tax_id, target_group) %>%
  count(target_group, name = "n") %>%
  mutate(target_group = fct_reorder(target_group, n)) %>%
  ggplot(aes(x = target_group, y = n, fill = target_group)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
    coord_flip() +
    labs(
      title = "RAMS Southern Ocean Whitelist — Taxa per Target Group",
      x     = "Target Group",
      y     = "Number of Taxa"
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 10))
```

```{r rams-target-groups-donut, echo=TRUE}
# Donut chart — proportional composition of RAMS whitelist target groups

rams_donut_data <- so_RAMS_list %>%
  distinct(tax_id, target_group) %>%
  count(target_group, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(
    pct          = n / sum(n),
    target_group = fct_reorder(target_group, n),
    label        = ifelse(
      pct * 100 >= 2,
      paste0(target_group, "\n", n,
             " (", scales::percent(pct, accuracy = 0.1), ")"),
      ""
    )
  )

ggplot(rams_donut_data, aes(x = 2, y = n, fill = target_group)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(
    aes(x = 2.6, label = label),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  xlim(0.5, 3.2) +
  labs(title = "RAMS Southern Ocean Whitelist — Target Group Composition") +
  theme_void() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))
```

```{r rams-target-groups-stacked, echo=TRUE}
# Stacked bar — class breakdown within each target group

so_RAMS_list %>%
  distinct(tax_id, target_group, class) %>%
  count(target_group, class) %>%
  mutate(target_group = fct_reorder(target_group, n, sum)) %>%
  ggplot(aes(x = target_group, y = n, fill = class)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "RAMS Whitelist — Class Breakdown within Target Groups",
      x     = "Target Group",
      y     = "Number of Taxa",
      fill  = "Class"
    ) +
    theme_minimal() +
    theme(legend.position = "none",
      axis.text.y     = element_text(size = 10),
      legend.text     = element_text(size = 7),
      legend.key.size = unit(0.4, "cm")
    )
```
### In the terminal/on server

<!-- ``` bash -->
<!-- conda install -c bioconda taxonkit -->
<!-- wget -c https://ftp.ncbi.nih.gov/pub/taxonomy/taxdump.tar.gz  -->
<!-- tar -zxvf taxdump.tar.gz -->

<!-- mkdir -p $HOME/.taxonkit -->
<!-- cp names.dmp nodes.dmp delnodes.dmp merged.dmp $HOME/.taxonkit -->

<!-- cut -f4 wgs_eukaryota.metadata_filter.short.tsv | tail -n+2 | taxonkit reformat2 - > taxon_paths.tsv -->
<!-- ``` -->

```{r join tables, echo=TRUE}

taxon_paths <- read_tsv("/Users/davisee/OneDrive - University of Tasmania/Documents/chapter2/raw data/taxon_paths.tsv", 
    col_names = c("tax_id", "taxonomic_path"),
    trim_ws = TRUE,
    col_types = cols(tax_id = col_character(),
  taxonomic_path = col_character()
  ))
wgs_euks2 <- wgs_euks |> inner_join(taxon_paths)


  
### Lets plot the reference metadata out first

# Bar chart of genome counts by source database
ggplot(wgs_euks2, aes(x = source_db)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3, size = 4) +
  labs(title = "", x = "", y = "Number of Genomes") +
  theme_minimal() 
```

```{r barplot-stacked, echo=TRUE}
# Stacked bar chart of assembly levels within each source database
ggplot(wgs_euks2, aes(x = source_db, fill = assembly_level)) +
  geom_bar(position = "stack") +
  labs(title = "", x = "Source Database", y = "Number of Genomes", fill = "Assembly Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot-proportional, echo=TRUE}
# Proportional stacked bar chart of assembly levels per source
wgs_euks2 %>% 
#  filter(is_SO==FALSE)%>%
ggplot(aes(x = source_db, fill = assembly_level)) +
  geom_bar(position = "fill") +
  labs(title = "proportional chart of all Genomes all assembly levels", x = "Source Database", y = "Proportion of Genomes", fill = "Assembly Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r apply-levels-assembly_level, echo = TRUE}
#to make DB filtering easier after understanding what types of assemblies are here, we are going to do some ranking.
unique(wgs_euks2$assembly_level)
#quality of assemblies goes "Chromosome", "complete Genome", "Contig", "Scaffold", so if we create these levels, we can understand white are the best assemblies either per accession, per species(tax_id) or both. This should signifcantly decrease database size. 

wgs_euks2$assembly_level <- factor(wgs_euks2$assembly_level, levels = c("Chromosome", "Complete Genome", "Contig", "Scaffold"))

levels(wgs_euks2$assembly_level)

```
```{r barplot-organelle-proportional, echo=TRUE}
#lets also explore what the scope of organelles is as well
# is there a skew of certain organelles at certain assembly levels? 
wgs_euks2 %>% 
ggplot(aes(x = organelle_description, fill = assembly_level)) +
  geom_bar(position = "fill") +
  labs(title = "proportional chart of all genome organelle type", x = "Organelle Description", y = "Proportion of Genomes", fill = "Assembly Level") +
  facet_wrap(~ source_db) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# so this is really good to know, as most of the NA's are probably nuclear data where they did not 

```{r barplot-SO genomes, echo=TRUE}
wgs_euks2 %>% 
  filter(is_SO==TRUE)%>%
ggplot(aes(x = source_db, fill = assembly_level)) + 
  geom_bar() +
   geom_text(stat = "count", 
             aes(label = after_stat(count)),
             size = 3, position = position_stack(vjust = 0.5)) + 
  labs(title = "Register of Marine Antarctic Species Found", x = "Source Database", y = "Number of Genomes", fill = "Assembly Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot-organelle-proportional, echo=TRUE}
#lets also explore what the scope of organelles is as well
#Proportional stacked bar chart of organelle type per source
wgs_euks2 %>%
  filter(is_SO==TRUE)%>%
ggplot(aes(x = source_db, fill = organelle_description)) +
  geom_bar(position = "fill") +
  labs(title = "Southern Ocean Taxa: proportional chart of all genome organelle type", x = "Source Database", y = "Proportion of Genomes", fill = "Orgenelle Description") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
## in these cases, NA usually refers to nuclear data 
```{r joining-tables, echo=TRUE}
#Let's review rarer organelle types to see if they would be good to keep in or not
unique(wgs_euks2$organelle_description)
#if I do filter, and it will be good to, would need to filter it by organelle_sequence_length, not ungapped_sequence_legnth & gapped_sequence_length
apico <- wgs_euks2 %>% filter(organelle_description=="Apicoplast") #apicoplasts might be good to keep bc pathogenitic groups, with cutoff 25,000 bp for organelle length
chloro <- wgs_euks2 %>% filter(organelle_description=="Chloroplast")
plast <- wgs_euks2 %>% filter(organelle_description=="Plastid") # also use 100,000 bp cutoff 
what <- wgs_euks2 %>% filter(organelle_description=="Mitochondrial Plasmid") #agricultural plasmid samples, REMOVE
misc <- wgs_euks2 %>% filter(organelle_description=="Mitochondrial Miscellaneous") # just four castor oil plant samples, REMOVE
mito <- wgs_euks2 %>% filter(organelle_description=="Mitochondrion")
chromo <- wgs_euks2 %>% filter(organelle_description == "Chromosome" | is.na(organelle_description))
# #remove the miscelleanous and mitochondrial plasmid after checking 
wgs_euks2 <- wgs_euks2 %>% 
  filter(!organelle_description %in% c("Mitochondrial Plasmid",
                                      "Mitochondrial Miscellaneous"))

#check to make sure misc filter worked properly, they should no longer be on the list! 
unique(wgs_euks2$organelle_description)
```
```{r filtering-organelle-dbs, echo = TRUE}
#check that each of the organelle vectors have the assembly_accsion levels that we applied earlier!
#per organelle, going to do the following steps
#(1)check that levels continued over for assembly_accession
#arrange by tax_id and then assembly acession, so for each taxa_id, assembly_acession goes first 
#run (distinct(tax_id, .keep_all=TRUE))

## ----------------- Apicoplast ----------------------- ##
levels(apico$assembly_level)

apico_slim <- apico %>%
  group_by(tax_id) %>%
  arrange(assembly_level, desc(genome_coverage)) %>%
  distinct(tax_id, .keep_all = TRUE) %>%
  ungroup()
#View(head(apico_slim, n = 5))

## ----------------- Chloroplast ---------------------- ##
levels(chloro$assembly_level)

chloro_slim <- chloro %>%
  group_by(tax_id) %>%
  arrange(assembly_level) %>%
  distinct(tax_id, .keep_all = TRUE)%>%
  ungroup()
#View(head(chloro_slim, n = 5))

## ------------------ Plastid ------------------------- ##
levels(plast$assembly_level)

plast_slim <- plast %>%
  group_by(tax_id) %>%
  arrange(assembly_level) %>%
  distinct(tax_id, .keep_all = TRUE)%>%
  ungroup()
#View(head(plast_slim, n = 5))

## ---------------- Mitochondrion --------------------- ##
levels(mito$assembly_level)

mito_slim <- mito %>%
  group_by(tax_id) %>%
  arrange(assembly_level) %>%
  distinct(tax_id, .keep_all = TRUE)%>%
  ungroup()
#View(head(mito_slim, n = 5))

## --------------- Chromosomes + NA (assumed nuclear) ------------ ##
levels(chromo$assembly_level)

chromo_slim <- chromo %>%
  group_by(tax_id) %>%
  arrange(assembly_level) %>%
  distinct(tax_id, .keep_all = TRUE)%>%
  ungroup()
#View(head(chromo_slim, n = 5))


```
```{r merge-back-db, echo=TRUE}
#merge all of the slim (filtered) db's back togethern to make to slim_wgs_euks2
############ --- merge all of them back together!!! ---- ########
wgs_euks2_slim <- bind_rows(chromo_slim, mito_slim, chloro_slim, plast_slim, apico_slim)

wgs_euks2_slim <- wgs_euks2_slim %>%
  mutate(assembly_level = as.character(assembly_level))

#check if the merge worked correctly 
wgs_euks2_slim %>% count(organelle_description)
```

```{r violin-plot, echo = TRUE}
## Review and filter out lower quality organelle assemblies ---> will then only catch complete organelles 
#Mitochondria Filter (most important for RAMS) organelle_length >= 10,000 bp 
##1 Check any that don't meet filter to see if they are Southern Ocean species
wgs_euks2_slim %>%
  filter(is_SO==FALSE) %>%
ggplot(aes(x = organelle_description, y = organelle_sequence_length, fill = assembly_level)) +
    geom_violin(alpha = 0.5, na.rm = TRUE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
    facet_wrap(~ source_db) +
    scale_y_continuous() +
    labs(
        title = "organelle length distribution by assembly Level",
        x = "Assembly Level",
        y = "Length bp", 
        fill = ""
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

### some oikopleura samples that are actually really important to keep in. so going to lower the cutoff t0 9000bp
```


```{r summary - SO scaffold genome exploration, echo=TRUE}
#isolate all all SO genomes at scaffold level to see what type of taxa they are
so_wgs_euks2_slim_subset <- wgs_euks2 %>% ##### MAKING A BIG CHANGE HERE
  filter(is_SO==TRUE) %>%
  mutate(assembly_level = as.character(assembly_level))
# it is a lot of important species, I think it would be good to keep them all in.

# length(unique(so_wgs_euks2_slim_subset$tax_id)) ## 143/ 558 SO taxa represented
# str_count(unique(so_wgs_euks2_slim_subset$tax_id)) ## each taxa represented by more than one genome 

noso_wgs_euks2_slim <- wgs_euks2_slim %>%
  filter(is_SO==FALSE) 

#Overall, would be good to keep in. 
# So, to make processing more distinct, let's split data going forward into two, the so_wgs_euks2_slim_subset, and then the filted_ccg_noSO_wgs_euks2_slim ----> wgs_euks3! 
```

``` {r joining-tables, echo = TRUE}
#filtering organelles#

# # 1) Mitochondrion: drop if length < 10,000 OR length is NA, but only for mito rows
# # 2) Chloroplast/Plastid: drop if length < 100,000 OR length is NA, only for those rows
# # 3) Apicoplast: drop if length < 30,000 OR length is NA, only for those rows



noso_wgs_euks2_slim_filtered <- noso_wgs_euks2_slim %>%
  mutate(
    keep = case_when(
      # Mitochondrion: require length >= 10,000 and not NA
      organelle_description == "Mitochondrion" ~
        !is.na(organelle_sequence_length) & organelle_sequence_length >= 10000,

      # Chloroplast or Plastid: require length >= 100,000 and not NA
      organelle_description %in% c("Chloroplast", "Plastid") ~
        !is.na(organelle_sequence_length) & organelle_sequence_length >= 100000,

      # Apicoplast: require length >= 30,000 and not NA
      organelle_description == "Apicoplast" ~
        !is.na(organelle_sequence_length) & organelle_sequence_length >= 30000,

      # Everything else (including organelle_description == NA): keep
      TRUE ~ TRUE
    )
  ) %>%
  filter(keep) %>%
  select(-keep)
```

```{r barplot-stacked, echo = TRUE}
## create new data with just assembly levels are most trustworthy after filtering by organelle length
# wgs_euks3 <- noso_wgs_euks2_slim_michap_filtered %>%
#   filter(assembly_level %in% c("Chromosome", "Complete Genome"))

# Define your level tiers
high_ok   <- c("Chromosome", "Complete Genome")
low_block <- c("Scaffold", "Contig")
#low_block <- c("Scaffold")

# 1) Collapse to *assembly-level decision*
by_asm <- noso_wgs_euks2_slim_filtered %>%
  filter(!is.na(assembly_accession), assembly_accession != "") %>%
  group_by(assembly_accession) %>%
  summarise(
    has_high = any(assembly_level %in% high_ok, na.rm = TRUE),
    has_low  = any(assembly_level %in% low_block, na.rm = TRUE),
    .groups  = "drop"
  )

# 2) Keep only assemblies that have high and do NOT have any low
asm_keep <- by_asm %>%
  filter(has_high, !has_low) %>%
  pull(assembly_accession)

# 3) From the original table, retain rows for those assemblies (if you still need rows)
wgs_euks3 <- noso_wgs_euks2_slim_filtered %>%
  filter(assembly_accession %in% asm_keep)

# 4) Immediately pick one representative row per assembly for downstream work
#    Prefer REFSEQ over GENBANK; break ties by longer lineage string
wgs_euks3_assemblies <- wgs_euks3 %>%
  arrange(desc(source_db == "REFSEQ"), desc(nchar(taxonomic_path))) %>%
  distinct(assembly_accession, .keep_all = TRUE)


# proportional Stacked bar chart of just highest assembly levels

wgs_euks3_assemblies %>% 
ggplot(aes(x = assembly_level, fill = source_db)) + 
  geom_bar() +
   geom_text(stat = "count", 
             aes(label = after_stat(count)),
             size = 3, position = position_stack(vjust = 0.5)) + 
  labs(title = "Chromosome & Complete Genome assembled Chordata Reference Genomes", x = "Assembly Level", y = "Number of Genomes", fill = "Source Database") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




```{r length_ratio, echo=TRUE}
# Step 1: Calculate length_ratio for both datasets 
wgs_euks3 <- wgs_euks3 %>%
  mutate(len_ratio = ifelse(
    is.na(ungapped_length) | is.na(sequence_length) | sequence_length == 0,
    NA_real_,
    ungapped_length / sequence_length
  ))

wgs_euks3_assemblies <- wgs_euks3_assemblies %>%
  mutate(len_ratio = ifelse(
    is.na(ungapped_length) | is.na(sequence_length) | sequence_length == 0,
    NA_real_,
    ungapped_length / sequence_length
  ))

# Step 2: Density plot colored by assembly_level and faceted by source_db
ggplot(wgs_euks3, aes(x = assembly_level, y = len_ratio, fill = assembly_level)) +
    geom_violin(alpha = 0.5, na.rm = TRUE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
    facet_wrap(~ source_db) +
    scale_y_continuous() +
    labs(
        title = "Length Ratio Distribution by high resolution Assembly Level",
        x = "Assembly Level",
        y = "Length Ratio"
    ) +
    theme_minimal() +
    theme(legend.position = "none")

```

```{r scatter-n50-l50, echo=TRUE}
# Scatter plot: Scaffold N50 vs Scaffold L50, faceted by source
ggplot(so_wgs_euks2_slim_subset, aes(x = scaffold_n50, y = scaffold_l50)) +
  geom_point(aes(size = contigs), alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(title = "RAMS taxa: All genomes no filtered", x = "Scaffold N50", y = "Scaffold L50") +
  theme_minimal()

#test
ggplot(wgs_euks3, aes(x = scaffold_n50, y = scaffold_l50)) +
  geom_point(aes(size = contigs), alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(title = "REfSeq and Genbank Chromosome + Complete Genomes, Organelle Filtered", x = "Scaffold N50", y = "Scaffold L50") +
  theme_minimal()


```

```{r scatter-n50-contigs-gc, echo=TRUE}
# Scatter plot: Scaffold N50 vs Contigs, filled by GC content
ggplot(so_wgs_euks2_slim_subset, aes(x = scaffold_n50, y = contigs)) +
  geom_point(aes(size = contigs, fill = gc_percent), shape = 21, color = "black", alpha = 0.8) +
  facet_wrap(~ source_db) +
  labs(title = "RAMS taxa: All genomes no filtered", x = "Scaffold N50", y = "Number of contigs", fill = "GC Percent") +
  theme_minimal()

ggplot(wgs_euks3, aes(x = scaffold_n50, y = contigs)) +
  geom_point(aes(size = contigs, fill = gc_percent), shape = 21, color = "black", alpha = 0.8) +
  facet_wrap(~ source_db) +
  labs(title = "REfSeq and Genbank Chromosome + Complete Genomes, Organelle Filtered", x = "Scaffold N50", y = "Number of contigs", fill = "GC Percent") +
  theme_minimal()
```
 
```{r genomes_lost_contigs, echo=TRUE}

# Define thresholds (descending from 500,000 in steps of 10,000)
thresholds <- seq(500000, 0, by = -100)

# For each source_db, count genomes <= threshold
threshold_by_source <- expand.grid(
  threshold = thresholds,
  source_db = unique(wgs_euks3$source_db)
) %>%
  rowwise() %>%
  mutate(n_remaining = sum(
    wgs_euks3$source_db == source_db &
      wgs_euks3$contigs <= threshold,
    na.rm = TRUE
  )) %>%
  ungroup()

# Make a plot by split databases
ggplot(threshold_by_source, aes(x = threshold, y = n_remaining, color = source_db)) +
  geom_point(size = 1) +
  scale_x_reverse(breaks = seq(0, 500000, 50000), labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Genome Retention by Contig Threshold, Split by Source DB",
    x = "Maximum Number of Contigs (Threshold)",
    y = "Number of Genomes Retained",
    color = "Source DB"
  ) +
  theme_minimal()


# Lets then merge the two sources and for each threshold, count genomes with contigs <= threshold
threshold_counts <- data.frame(threshold = thresholds) %>%
  rowwise() %>%
  mutate(n_remaining = sum(wgs_euks3$contigs <= threshold, na.rm = TRUE)) %>%
  ungroup()

# Reverse x and y for proper elbow detection
x <- rev(threshold_counts$threshold)
y <- rev(threshold_counts$n_remaining)

# Detect elbow (tipping point)
elbow_index <- inflection::findiplist(x, y, index = 1)
tipping_threshold <- x[elbow_index]
n_at_tipping <- y[elbow_index]

# Plot with tipping point highlighted
ggplot(threshold_counts, aes(x = threshold, y = n_remaining)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line(color = "steelblue") +
  geom_vline(xintercept = tipping_threshold, linetype = "dashed", color = "red") +
  geom_text(
    data = data.frame(
      threshold = tipping_threshold,
      n_remaining = n_at_tipping,
      label = paste0("Tipping Point: ", tipping_threshold, " contigs")
    ),
    aes(x = threshold, y = n_remaining, label = label),
    inherit.aes = FALSE,
    vjust = -1, hjust = 1.1, color = "red"
  ) +
  scale_x_reverse(breaks = seq(0, 500000, 50000), labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Tipping Point in Genome Retention Curve",
    x = "Contig Threshold",
    y = "Number of Genomes Retained"
  ) +
  theme_minimal()

elbows <- inflection::findiplist(x, y, index = 1)
tipping_thresholds <- x[elbows]

con_thres <- max(tipping_thresholds, na.rm = TRUE)


```


```{r density-gc, echo=TRUE}
# Density plot of GC percent by assembly level
ggplot(wgs_euks3, aes(x = as.numeric(gc_percent), fill = assembly_level)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ source_db) +
  labs(title = "REfSeq and Genbank Chromosome + Complete Genomes, Organelle Filtered", x = "GC Percent", y = "Density", fill = "Assembly Level") +
  theme_minimal()
ggplot(so_wgs_euks2_slim_subset, aes(x = as.numeric(gc_percent), fill = assembly_level)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ source_db) +
  labs(title = "RAMS taxa: All genomes no filtered", x = "GC Percent", y = "Density", fill = "Assembly Level") +
  theme_minimal()
```

```{r ungapped-vs-total, echo=TRUE}
# Scatter plot comparing ungapped and total sequence lengths
ggplot(wgs_euks3, aes(x = ungapped_length, y = sequence_length)) +
  geom_point(alpha = 0.8, aes(color = assembly_level)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
  facet_wrap(~ source_db) +
  labs(title = "REfSeq and Genbank Chromosome + Complete Genomes, Organelle Filtered", x = "Ungapped Sequence Length", y = "Total Sequence Length") +
  theme_minimal()
ggplot(so_wgs_euks2_slim_subset, aes(x = ungapped_length, y = sequence_length)) +
  geom_point(alpha = 0.8, aes(color = assembly_level)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
  facet_wrap(~ source_db) +
  labs(title = "RAMS taxa: All genomes no filtered", x = "Ungapped Sequence Length", y = "Total Sequence Length") +
  theme_minimal()


```


```{r ungapped-vs-total-ratio, echo=TRUE}
# how many sequence are 10% longer than their ungapped sequence length?

wgs_euks3 %>% filter(source_db == "GENBANK") |> 
  filter(sequence_length / ungapped_length  > 1.1) %>% 
  nrow()
wgs_euks3_assemblies %>% filter(source_db == "GENBANK") |> 
  filter(sequence_length / ungapped_length  > 1.1) %>% 
  nrow()
# Let's add a length_ratio to the dataframe
wgs_euks3 <- wgs_euks3 %>%
  mutate(len_ratio = (sequence_length / ungapped_length))
summary(wgs_euks3$len_ratio)
wgs_euks3_assemblies <- wgs_euks3_assemblies %>%
  mutate(len_ratio = (sequence_length / ungapped_length))


#add length ratio to southern ocean ones
so_wgs_euks2_slim_subset <- so_wgs_euks2_slim_subset %>%
  mutate(len_ratio = (sequence_length / ungapped_length))


summary(wgs_euks3_assemblies$len_ratio)
len_ratio <- 1.1 
```

### plots the points filtered by the len_ratio threshold

```{r plot-len_ratio-boxplot, echo=TRUE}
# Step 2: Density plot colored by assembly_level and faceted by source_db
ggplot(wgs_euks3_assemblies, aes(x = assembly_level, y = len_ratio, fill = assembly_level)) +
    geom_violin(alpha = 0.5, na.rm = TRUE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
    geom_jitter() +
    facet_wrap(~ source_db) +
    scale_y_continuous() +
    labs(
        title = "Length Ratio Distribution by Assembly Level",
        x = "Assembly Level",
        y = "How much longer is the total sequence length vs the ungapped length"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r plot-len_ratio-filter, echo=TRUE}
ggplot(wgs_euks3_assemblies, aes(x = ungapped_length, y = sequence_length)) +
  geom_point(alpha = 0.8, aes(color = assembly_level)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
  facet_wrap(~ source_db, drop = FALSE) +  # This keeps all source_db panels
  gghighlight(len_ratio > 1.1, use_direct_label = FALSE) +
  labs(
    title = "Ungapped vs Total Sequence Length (Highlighting len_ratio > 1.1)",
    x = "Ungapped Sequence Length",
    y = "Total Sequence Length"
  ) +
  theme_minimal()

```

### plots the points filtered by the len_ratio threshold and that are register Antarctic Marine species (RAMS list - Chordata) 
# ```{r plot-len_ratio-filter-iridian, echo=TRUE}
# ggplot(wgs_euks2_slim, aes(x = ungapped_length, y = sequence_length)) +
#   geom_point(alpha = 0.8, aes(color = assembly_level)) +
#   geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "lightgrey") +
#   facet_wrap(~ source_db, drop = FALSE) +
#   gghighlight(is_SO == TRUE, use_direct_label = FALSE) +
#   labs(
#     title = "Ungapped vs Total Sequence Length (Highlighting Iridian Genomes)",
#     x = "Ungapped Sequence Length",
#     y = "Total Sequence Length"
#   ) +
#   theme_minimal()

<!-- ``` -->


### We found quite a bit of NAs in the scaffold_n50, lets see which assembly level they are
### This will be good to check for both DB, will it still be too large? 

#write.csv(wgs_euks2_slim, file ="wgs_euks2_slim.csv")

```{r counting-nas, echo=TRUE}
# # Count NA values per column and source_db
# na_counts <- wgs_euks3 %>%
#   select(source_db, scaffold_n50, contigs) %>%
#   pivot_longer(cols = c(scaffold_n50, contigs), names_to = "variable", values_to = "value") %>%
#   mutate(is_na = is.na(value)) %>%
#   group_by(source_db, variable) %>%
#   summarise(n = sum(is_na), .groups = "drop") %>%
#   complete(source_db, variable, fill = list(n = 0))
# 
# # Plot grouped by source_db, faceted by column
# ggplot(na_counts, aes(x = source_db, y = n, fill = source_db)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~ variable, scales = "free_y") +
#   labs(title = "Number of NAs by Column and Source DB",
#        x = "Source DB",
#        y = "Number of NAs") +
#   theme_minimal()
# 
# 
# # Count NAs in scaffold_n50 and show their assembly_levels
# na_pie_data <- wgs_euks3   %>%
#   filter(is.na(scaffold_n50)) %>%
#   count(assembly_level) %>%
#   mutate(percentage = n / sum(n),
#          label = paste0(assembly_level, " (", round(percentage * 100, 1), "%)"))
# 
# ggplot(na_pie_data, aes(x = "", y = n, fill = assembly_level)) +
#   geom_col(width = 1, color = "white") +
#   coord_polar(theta = "y") +
#   geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
#   labs(title = "Distribution of NAs in scaffold_n50 by Assembly Level",
#        x = NULL, y = NULL, fill = "Assembly Level") +
#   theme_void()
```

Doing a manual check on these the genomes looks ok, one option could be to filter these based on the coverage they have. We observe a suspiscious decline in the lower end of the coverage. This high value is caused by the extreme high values and hence skewness in the data. We therefore selected a hard thresholds of 10. For setting the contigs we use the .01 / .99 percentile of RefSeq stats to filter the Genbank genomes. as well as the length_ratio


One idea is to use the RefSeq as a gold standard, and use the scaffold_n50 and contigs statistics to filter the genbank.


Elisa: for my stuff, this standard above could be good to filter SO_genomes list, but then to a hard cut-off of taking out all scaffold level assemblies, need to have a think to set contig-level assembly filters (more stringent length_ratio, coverage, ATGC percentage). 

For So_genomes, need to probably do in depth based R markdown on included scaffold 
(following these derived stringent thresholds and stuff. So basically, should create a new version + 2nd part to do SO genome curation analysis and the larger one.)

```{r scaffols_n50_NA_contig_genome_coverage, echo=TRUE}

# Setting thresholds
scaf_thres <- quantile(wgs_euks$scaffold_n50[wgs_euks$source_db == "REFSEQ"], 0.01, na.rm = TRUE)
con_thres <- quantile(wgs_euks$contigs[wgs_euks$source_db == "REFSEQ"], 0.99, na.rm = TRUE) # Is now set as the tipping point from the above elbow plot
gen_cov_thres <- 25
len_thres <- 1.1

# Count how many will be filtered by this 
highlighted_counts <- wgs_euks3_assemblies %>%
  filter(
    !(
      (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
      (is.na(scaffold_n50)    | scaffold_n50 >= scaf_thres) &
      (is.na(contigs)         | contigs <= con_thres) &
      (is.na(len_ratio)         | len_ratio <= len_thres)
    )
  ) %>%
  group_by(source_db) %>%
  summarise(count = n())


ggplot(wgs_euks3_assemblies, aes(x = genome_coverage, y = contigs)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  scale_x_log10() +
  labs(
    title = "",
    x = "genome_coverage",
    y = "contigs"
  ) +
  gghighlight(scaffold_n50 > scaf_thres, contigs < con_thres, genome_coverage >= gen_cov_thres, use_direct_label = FALSE) +
  geom_text(
    data = highlighted_counts,
    aes(x = Inf, y = Inf, label = paste("To be filtered:", count)),
    hjust = 1.1, vjust = 1.5,
    inherit.aes = FALSE,
    size = 4,
    fontface = "bold",
    color = "darkgrey"
  ) +
  theme_minimal()
```

```{r scatter-contigs-n50, echo=TRUE}
# Contigs vs Scaffold N50
# Filter input for plotting (remove NA and zero values for valid plotting)
plot_data <- wgs_euks3_assemblies %>%
  filter(!is.na(scaffold_n50), !is.na(contigs), scaffold_n50 > 0, contigs > 0)

# Step 1: Filter and count per source_db (MATCH gghighlight thresholds)
filtered_counts <- wgs_euks3_assemblies %>% filter(
    (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
    (is.na(scaffold_n50) | scaffold_n50 >= scaf_thres) &
    (is.na(contigs) | contigs <= con_thres) &
      (is.na(len_ratio)         | len_ratio <= len_thres)
  ) %>%
  count(source_db, name = "filtered_n")

# Step 2: Compute label positions (adjust x/y to avoid overlap)
label_positions <- plot_data %>%
  group_by(source_db) %>%
  summarize(
    x = quantile(contigs, 0.05, na.rm = TRUE),
    y = max(scaffold_n50, na.rm = TRUE) * 0.95,
    .groups = "drop"
  )

# Step 3: Join label data
label_data <- left_join(filtered_counts, label_positions, by = "source_db")

# Step 4: Create the plot
ggplot(plot_data, aes(x = contigs, y = scaffold_n50)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ source_db) +
  labs(
    title = "Filtered Genome Count per Source DB",
    x = "Number of Contigs",
    y = "Scaffold N50"
  ) +
  theme_minimal() +
  gghighlight(scaffold_n50 > scaf_thres, contigs < con_thres,genome_coverage >= gen_cov_thres, use_direct_label = FALSE) +
  scale_x_continuous(labels = scales::label_comma()) +
  geom_text(
    data = label_data,
    aes(x = x, y = y, label = paste("Kept genomes:", scales::comma(filtered_n))),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0,
    fontface = "bold"
  )
```


```{r filter-and-flag-large, echo=TRUE}
# Apply quality thresholds to non-SO genomes
# Always keep organelles; apply coverage/contig/length thresholds to nuclear assemblies

org_whitelist <- c("Mitochondrion", "Apicoplast", "Plastid")

wgs_euks3_flagged_large <- wgs_euks3 %>%
  mutate(
    keep = case_when(
      organelle_description %in% org_whitelist ~ TRUE,
      is.na(organelle_description) | organelle_description %in% c("Chromosome", "Chloroplast") ~
        (is.na(genome_coverage) | genome_coverage >= gen_cov_thres) &
        (is.na(scaffold_n50)    | scaffold_n50    >= scaf_thres)    &
        (is.na(contigs)         | contigs         <= con_thres)     &
        (is.na(len_ratio)       | len_ratio       <= len_thres),
      TRUE ~ FALSE
    )
  )
```

```{r pie-filter-status, echo=TRUE}
# Pie charts showing filter status (in vs out) per source database
library(purrr)
library(patchwork)

pie_data <- wgs_euks3_flagged_large %>%
  mutate(filter_status = ifelse(keep, "Filtered In", "Filtered Out")) %>%
  count(source_db, filter_status) %>%
  group_by(source_db) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label   = paste0(filter_status, " ", percent, "%")
  ) %>%
  ungroup()

split_pie_data <- split(pie_data, pie_data$source_db)

pie_plots <- map(split_pie_data, function(df) {
  ggplot(df, aes(x = "", y = n, fill = filter_status)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
    scale_fill_manual(values = c("Filtered In" = "steelblue", "Filtered Out" = "grey70")) +
    labs(title = paste("Filter Status for", unique(df$source_db))) +
    theme_void()
})

wrap_plots(pie_plots)
```

```{r split-taxonomic-ranks, echo=TRUE}
# Split taxonomic path into rank columns for both non-SO and SO subsets

wgs_euks3_flagged_split <- wgs_euks3_flagged_large %>%
  separate(
    taxonomic_path,
    into = paste0("rank", 1:7),
    sep  = ";",
    fill = "right"
  ) %>%
  mutate(across(starts_with("rank"), trimws)) %>%
  filter(rank4 != "Primates" | is.na(rank4))   # remove Primates except one that will be added in with stuff later

########### lets also look at split in non-target
wgs_euks3_flagged_split %>%
  distinct(assembly_accession, rank2, rank3) %>%
  count(rank2, rank3) %>%
  arrange(desc(n)) %>%
  print(n = Inf)

# Apply same split to SO genomes (use wgs_euks2 joined version)
so_wgs_euks3_split <- so_wgs_euks2_slim_subset %>%
  separate(
    taxonomic_path,
    into = paste0("rank", 1:7),
    sep  = ";",
    fill = "right"
  ) %>%
  mutate(across(starts_with("rank"), trimws))

# Review what is the actual taxonomic breakdown of available 
so_wgs_euks3_split %>%
  distinct(assembly_accession, rank2, rank3) %>%
  count(rank2, rank3) %>%
  arrange(desc(n)) %>%
  print(n = Inf)
```

```{r barplot-filtered-out-by-group, echo=TRUE}
# Bar chart of filtered-out genomes by eukaryotic group and assembly level

library(forcats)

remove_data <- wgs_euks3_flagged_split %>%
  filter(!keep) %>%
  count(rank2, source_db, assembly_level, name = "n") %>%
  mutate(rank2 = fct_reorder(rank2, n))

ggplot(remove_data, aes(x = rank2, y = n, fill = assembly_level)) +
  geom_col() +
  facet_wrap(~ source_db, scales = "free_y") +
  labs(
    title = "Filtered Out Eukaryotic Groups by Source DB and Assembly Level",
    x     = "Eukaryotic Group",
    y     = "Number of Genomes",
    fill  = "Assembly Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r split-keep-remove, echo=TRUE}
# Split into kept and removed subsets for GENBANK and REFSEQ

genbank_remove <- wgs_euks3_flagged_split %>% filter(source_db == "GENBANK", !keep)
genbank_keep   <- wgs_euks3_flagged_split %>% filter(source_db == "GENBANK",  keep)
refseq_remove  <- wgs_euks3_flagged_split %>% filter(source_db == "REFSEQ",  !keep)
refseq_keep    <- wgs_euks3_flagged_split %>% filter(source_db == "REFSEQ",   keep)

all_remove <- wgs_euks3_flagged_split %>%
  filter(source_db %in% c("GENBANK", "REFSEQ"), !keep)
all_keep   <- wgs_euks3_flagged_split %>%
  filter(source_db %in% c("GENBANK", "REFSEQ"),  keep)

# Merge kept non-SO genomes with all SO genomes
all_keep_final <- bind_rows(all_keep, so_wgs_euks3_split)
```

```{r save-outputs, echo=TRUE}
# Save kept, removed and SO genome tables

write.csv(all_remove,        file = "removed_genomes_nocontig.csv")
write.csv(all_keep,          file = "kept_genomes_nocontig.csv")
write.csv(so_wgs_euks3_split, file = "so_kept_genomes.csv")
```

```{r review-all-keep, echo=TRUE}
# Review unique taxa and assembly representation in all_keep_final

skinny_all_keep <- all_keep_final %>%
  distinct(assembly_accession, tax_id)

test_skinny_all_keep <- all_keep_final %>%
  distinct(tax_id)
cat("Unique tax_ids in all_keep_final:", nrow(test_skinny_all_keep), "\n")

test_taxid_assembly_all_keep <- all_keep_final %>%
  distinct(tax_id, assembly_level)

test_organelle_assembly_all_keep <- all_keep_final %>%
  distinct(tax_id, organelle_description)
```

```{r merge-and-export-accessions, echo=TRUE}
# Combine SO and non-SO kept genomes and export assembly accession lists

library(readr)

merged_tbl <- bind_rows(all_keep, so_wgs_euks2_slim_subset)

assembly_ids <- merged_tbl %>%
  distinct(assembly_accession) %>%
  arrange(assembly_accession)
```
```{save-outputs, echo = TRUE}
write_csv(assembly_ids, "assembly_accessions_nocontig_SLIM.csv")
write_lines(assembly_ids$assembly_accession, "assembly_accessions_SLIM_nocontig.txt")
```

```{r assign-taxonomic-groups, echo=TRUE}
# Assign broad taxonomic groups to each assembly for downstream analysis

assign_db_group <- function(rank2, rank3) {
  case_when(
    rank2 == "Chordata" & rank3 %in% c("Actinopteri", "Chondrichthyes",
                                        "Cladistia", "Hyperoartia", "Myxini") ~ "Fish",
    rank2 == "Chordata" & rank3 == "Mammalia"                                 ~ "Mammalia",
    rank2 == "Chordata" & rank3 == "Aves"                                     ~ "Birds",
    rank2 == "Chordata" & rank3 == "Lepidosauria"                             ~ "Reptiles",
    rank2 == "Chordata" & rank3 == "Amphibia"                                 ~ "Amphibians",
    rank2 %in% c("Chordata", "Hemichordata")                                  ~ "Chordates_Outgroup",
    grepl("Arthropoda|Mollusca|Echinodermata|Nematoda|Platyhelminthes|Annelida|Porifera|Cnidaria",
          rank2, ignore.case = TRUE)                                           ~ "Invertebrates",
    grepl("Ascomycota|Basidiomycota|Mucoromycota|Chytridiomycota|Microsporidia",
          rank2, ignore.case = TRUE)                                           ~ "Fungi_Protists",
    TRUE                                                                       ~ "Plants_Algae_Other"
  )
}

accessions_with_groups <- merged_tbl %>%
  mutate(db_group = assign_db_group(rank2, rank3)) %>%
  distinct(tax_id, db_group, rank2, rank3, rank4, rank5, rank6) %>%
  arrange(db_group, tax_id)

accessions_with_groups %>%
  count(db_group, name = "n_accessions") %>%
  arrange(desc(n_accessions)) %>%
  print(n = Inf)

write_csv(accessions_with_groups, "assembly_accessions_with_groups_extra_taxonomic.csv")
message("Done. Taxonomic group accession table written.")
```

```{r scatter-chordata-SO, echo=TRUE}
# Contigs vs Scaffold N50 highlighting Southern Ocean Chordata genomes

merged_tbl %>%
  filter(rank2 == "Chordata") %>%
  ggplot(aes(x = contigs, y = scaffold_n50, shape = is_SO)) +
    geom_point(alpha = 0.7) +
    facet_wrap(~ source_db) +
    labs(
      title = "Chordata Genomes — SO taxa highlighted",
      x     = "Number of Contigs",
      y     = "Scaffold N50",
      shape = "SO Genome"
    ) +
    scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
    scale_x_continuous(labels = scales::label_comma()) +
    theme_minimal() +
    gghighlight(is_SO == TRUE, use_direct_label = FALSE)
```

```{r scatter-streptophyta-SO, echo=TRUE}
# Contigs vs Scaffold N50 for Streptophyta (plants) highlighting SO taxa

genbank_keep %>%
  filter(rank2 == "Streptophyta") %>%
  ggplot(aes(x = contigs, y = scaffold_n50, shape = is_SO)) +
    geom_point(alpha = 0.7) +
    facet_wrap(~ source_db) +
    labs(
      title = "Streptophyta (Plant) Genomes — SO taxa highlighted",
      x     = "Number of Contigs",
      y     = "Scaffold N50",
      shape = "SO Genome"
    ) +
    scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
    scale_x_continuous(labels = scales::label_comma()) +
    theme_minimal() +
    gghighlight(is_SO == TRUE, use_direct_label = FALSE)
```

```{r donut-taxonomy, echo=TRUE}
# Donut charts showing taxonomic composition of merged_tbl at multiple ranks

library(scales)
library(forcats)

prep_pie_data <- function(df, col, na_label = "Unknown", label_min_pct = 2) {
  col <- rlang::enquo(col)
  df %>%
    transmute(level = fct_na_value_to_level(factor(!!col), level = na_label)) %>%
    count(level, name = "n") %>%
    arrange(desc(n)) %>%
    mutate(
      pct   = n / sum(n),
      level = fct_reorder(level, n),
      label = paste0(level, " — ", n),
      label = ifelse(pct * 100 >= label_min_pct, label, "")
    )
}

plot_donut <- function(pie_df, title) {
  ggplot(pie_df, aes(x = 2, y = n, fill = level)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    geom_text(
      aes(x = 2.05, label = label),
      position = position_stack(vjust = 0.5),
      size = 3
    ) +
    xlim(0.5, 2.6) +
    labs(title = title) +
    theme_void() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
}

rank2_pie_df <- prep_pie_data(merged_tbl, rank2, label_min_pct = 2)

strep_rank4_pie_df <- merged_tbl %>%
  filter(rank2 == "Streptophyta") %>%
  prep_pie_data(rank4, label_min_pct = 4)

arthro_rank4_pie_df <- merged_tbl %>%
  filter(rank2 == "Arthropoda") %>%
  prep_pie_data(rank4, label_min_pct = 4)

strep_rank6_pie_df <- merged_tbl %>%
  filter(rank2 == "Streptophyta") %>%
  prep_pie_data(rank6, label_min_pct = 4)

donut_rank2       <- plot_donut(rank2_pie_df,       "All taxa — Phylum (rank2)")
donut_strep_rank4 <- plot_donut(strep_rank4_pie_df, "Streptophyta — Orders")
donut_arthro_rank4 <- plot_donut(arthro_rank4_pie_df, "Arthropoda — Orders")
donut_strep_rank6 <- plot_donut(strep_rank6_pie_df, "Streptophyta — Genus (rank6)")

donut_rank2 | donut_arthro_rank4 | donut_strep_rank4 
```



```{r so-target-groups, echo=TRUE}
# Assign target groups to so_wgs_euks3_split based on rank2 and rank3

so_wgs_euks3_split <- so_wgs_euks3_split %>%
  mutate(
    target_group = case_when(
      rank2 == "Chordata" & rank3 == "Actinopteri"    ~ "Fish",
      rank2 == "Chordata" & rank3 == "Chondrichthyes" ~ "Fish",
      rank2 == "Chordata" & rank3 == "Hyperoartia"    ~ "Fish",
      rank2 == "Chordata" & rank3 == "Mammalia"       ~ "Mammalia",
      rank2 == "Chordata" & rank3 == "Aves"           ~ "Birds",
      rank2 == "Chordata" & rank3 %in% c("Ascidiacea",
                                          "Thaliacea") ~ "Tunicata",
      rank2 == "Mollusca"                              ~ "Mollusca",
       rank2 == "Mollusca"                              ~ "Mollusca",
      rank2 == "Arthropoda" & rank3 %in% c("Malacostraca",
                                           	"Ostracoda", 
                                           "Hexactinellida")~ "Crustaceans",
      rank2 == "Arthropoda" & rank3 =="Pycnogonida"  ~ "SeaSpiders",
      rank2 == "Echinodermata"                         ~ "Echinodermata",
      rank2 == "Porifera"                              ~ "Porifera",
      rank2 == "Annelida"                              ~ "Annelida",
      rank2 == "Cnidaria"                              ~ "Cnidaria",
      rank2 == "Bryozoa"                               ~ "Bryozoa",
      rank2 == "Ctenophora"                            ~ "Ctenophora",
      rank2 == "Nematoda"                              ~ "Nematoda",
      TRUE                                             ~ "Other"
    )
  )

cat("SO target group breakdown:\n")
print(table(so_wgs_euks3_split$target_group, useNA = "always"))

################ non - southern ocean 
wgs_euks3_flagged_split <- wgs_euks3_flagged_split %>%
  mutate(
    target_group = case_when(
      rank2 == "Chordata" & rank3 == "Actinopteri"    ~ "Fish",
      rank2 == "Chordata" & rank3 == "Chondrichthyes" ~ "Fish",
      rank2 == "Chordata" & rank3 == "Hyperoartia"    ~ "Fish",
      rank2 == "Chordata" & rank3 == "Mammalia"       ~ "Mammalia",
      rank2 == "Chordata" & rank3 == "Aves"           ~ "Birds",
      rank2 == "Chordata" & rank3 %in% c("Ascidiacea",
                                          "Thaliacea") ~ "Tunicata",
      rank2 == "Mollusca"                              ~ "Mollusca",
      rank2 == "Arthropoda" & rank3 %in% c("Malacostraca",
                                           	"Ostracoda", 
                                           "Hexactinellida")~ "Crustaceans",
      rank2 == "Arthropoda" & rank3 =="Pycnogonida"  ~ "SeaSpiders",
      rank2 == "Echinodermata"                         ~ "Echinodermata",
      rank2 == "Porifera"                              ~ "Porifera",
      rank2 == "Annelida"                              ~ "Annelida",
      rank2 == "Cnidaria"                              ~ "Cnidaria",
      rank2 == "Bryozoa"                               ~ "Bryozoa",
      rank2 == "Ctenophora"                            ~ "Ctenophora",
      rank2 == "Nematoda"                              ~ "Nematoda",
      TRUE                                             ~ "Other"
    )
  )

cat("Non-SO target group breakdown:\n")
print(table(wgs_euks3_flagged_split$target_group, useNA = "always"))




```

```{r so-target-groups-barplot, echo=TRUE}
# Bar chart — unique assemblies per target group coloured by assembly level

so_wgs_euks3_split %>%
  distinct(assembly_accession, target_group, assembly_level, source_db) %>%
  count(target_group, assembly_level, source_db) %>%
  mutate(target_group = fct_reorder(target_group, n, sum)) %>%
  ggplot(aes(x = target_group, y = n, fill = assembly_level)) +
    geom_col() +
    facet_wrap(~ source_db, scales = "free_y") +
    labs(
      title = "Southern Ocean Genomes by Target Group",
      x     = "Target Group",
      y     = "Number of Assemblies",
      fill  = "Assembly Level"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r so-target-groups-donut, echo=TRUE}
# Donut chart — proportional composition of SO target groups

so_donut_data <- so_wgs_euks3_split %>%
  distinct(tax_id, target_group) %>%
  count(target_group, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(
    pct          = n / sum(n),
    target_group = fct_reorder(target_group, n),
    label        = ifelse(
      pct * 100 >= 2,
      paste0(target_group, " ", " (",n , " - ", scales::percent(pct, accuracy = 0.1), ")"),
      ""
    )
  )

ggplot(so_donut_data, aes(x = 2, y = n, fill = target_group)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(
    aes(x = 2.7, label = label),
    position = position_stack(vjust = 0.3),
    size = 3
  ) +
  xlim(0.5, 3.2) +
  labs(title = "Southern Ocean Genomes — Target Group Composition") +
  theme_void() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```